---
# ==========================================================
# Personal Economic Research Portfolio
# ==========================================================

title: "The Economics of Human Shrinkage"
subtitle: "Economic Drivers of Global Fertility Contraction: A Panel Fixed-Effects Analysis"

author:
  - name: "Abzhamiev Bekdaulet"
    affiliation: "Independent Economic Researcher"
    orcid: "0009-0003-4999-1861"

date: today
date-format: "D MMMM YYYY"
lang: en

header-includes:
  - \orcid{0009-0003-4999-1861}

# ==========================================================
---

```{r}
#| include: false

source("libraries.R")
```

\newpage

# Introduction and Hypotheses

Modern demography is characterized by a steady decline in the **Total Fertility Rate (TFR)** globally. In advanced economies, TFR is consistently below the replacement level ($\approx 2.1$). This study evaluates the structural socio-economic determinants of fertility using World Bank panel data from 1990 to 2024.

We test **five distinct hypotheses**: \* $H_1$ (Urbanization): Higher urbanization rates correlate with lower TFR. \* $H_2$ (Income): Higher income per capita correlates with lower TFR. \* $H_3$ (Female Labor): Higher female labor force participation correlates with lower TFR. \* $H_4$ (Life Expectancy): Higher life expectancy correlates with lower TFR. \* $H_5$ (Infant Mortality): Higher infant mortality rates correlate with higher TFR.

## Microeconomic Formalization & Quantity-Quality Trade-off Model

To formalize the economic drivers of fertility contraction, we extend Becker's (1960) microeconomic household production model. A rational household maximizes a utility function defined over consumption goods ($C$), the number of children ($N$), and the quality/human-capital investment per child ($Q$):

$$U = U(C, N, Q) = C^{\alpha} N^{\beta} Q^{\gamma}, \quad \text{where } \alpha + \beta + \gamma = 1$$

Subject to the comprehensive household budget constraint:

$$I = C + P_n \cdot N + P_q \cdot N \cdot Q + w \cdot \tau \cdot N$$

Where:
* $I$ represents total household potential income.
* $P_n$ is the baseline maintenance cost per child (food, basic shelter).
* $P_q$ is the price of human capital investment (education, healthcare).
* $w \cdot \tau$ reflects the opportunity cost of maternal time ($\tau$) allocated to child-rearing at market wage rate $w$ (directly capturing Female Labor Force Participation effects).

### Dynamic Panel Econometric Specification

To account for intertemporal demographic momentum and path dependency, we specify the empirical dynamic Panel Two-Way Fixed Effects model:

$$TFR_{i,t} = \delta TFR_{i,t-1} + \beta_1 \text{Urban}_{i,t} + \beta_2 \ln(\text{Income}_{i,t}) + \beta_3 \text{FLFP}_{i,t} + \beta_4 \text{LifeExp}_{i,t} + \beta_5 \text{InfantMort}_{i,t} + \mu_i + \lambda_t + \varepsilon_{i,t}$$

Where $\mu_i$ captures unobserved time-invariant country heterogeneity, $\lambda_t$ absorbs global macroeconomic/demographic time shocks, and $\varepsilon_{i,t}$ is the idiosyncratic error term evaluated using Arellano-Bond Generalized Method of Moments (GMM) robust estimators to prevent endogeneity biases.

# Environment Setup and Library Loading

First, we load the required packages for data manipulation, API queries, and visualization.

```{r}
#| label: setup-libraries
#| echo: false
#| message: false
#| warning: false

# Автоматическая установка недостающих пакетов
required_packages <- c("tidyverse", "wbstats", "gridExtra")
new_packages <- required_packages[!(required_packages %in% installed.packages()[,"Package"])]
if(length(new_packages)) install.packages(new_packages)

# Подключение библиотек
library(tidyverse)
library(wbstats)
library(gridExtra)
```

# Data Acquisition via World Bank API

We extract global panel data programmatically using the `wbstats` API connector. We target only actual countries (excluding regional aggregates) and clear missing values to ensure statistical validity.

```{r}
#| label: load-and-clean-data
#| echo: false
#| message: false
#| warning: false

# Запрос к API Всемирного банка
raw_demographics <- wb_data(
  country = "countries_only", # Скачиваем данные только по чистым странам
  indicator = c(
    "SP.DYN.TFRT.IN",     # TFR
    "SP.URB.TOTL.IN.ZS",   # Urbanization (%)
    "NY.GDP.PCAP.KD",     # GDP per Capita
    "SL.TLF.CACT.FE.ZS",   # Female Labor Force Participation (%)
    "SP.DYN.LE00.IN",     # Life Expectancy
    "SP.DYN.IMRT.IN"      # Infant Mortality
  ),
  start_date = 1990,
  end_date = 2024
)

# Очистка, переименование и фильтрация данных с явными вызовами dplyr::
panel_data <- raw_demographics %>%
  dplyr::rename(
    year = date,
    tfr = SP.DYN.TFRT.IN,
    urban = SP.URB.TOTL.IN.ZS,
    income = NY.GDP.PCAP.KD,
    female_labor = SL.TLF.CACT.FE.ZS,
    life_exp = SP.DYN.LE00.IN,
    infant_mort = SP.DYN.IMRT.IN
  ) %>%
  dplyr::select(country, year, tfr, urban, income, female_labor, life_exp, infant_mort) %>%
  drop_na() # Удаляем пропущенные наблюдения

# Отображение первых 5 строк полученной панели
knitr::kable(head(panel_data, 5), digits = 2, caption = "Preview of the Cleared Global Panel Data")
```

# Statistical Testing: Correlation Matrix

To evaluate the direction and strength of the relationships for all five hypotheses simultaneously, we construct a Pearson correlation matrix using our filtered global panel dataset.

```{r}
#| label: correlation-matrix
#| echo: false
#| message: false
#| warning: false

# Вычисляем корреляцию, явно вызывая dplyr::select
correlation_results <- cor(panel_data %>% dplyr::select(-country, -year))

# Строим аккуратную таблицу корреляций
knitr::kable(correlation_results, digits = 2, caption = "Global Correlation Matrix (1990-2024)")
```

# Visualizing Demographic Shifts (H1 and H4)

To visually confirm our highest statistical correlations, we plot the relationships between the Total Fertility Rate (TFR) and both the Urbanization Rate ($H_1$) and Life Expectancy ($H_4$) across our global panel.

```{r}
#| label: trend-plots-h1-h4
#| fig-width: 12
#| fig-height: 5
#| echo: false

# График H1: Урбанизация против TFR
plot_h1 <- ggplot(panel_data, aes(x = urban, y = tfr)) +
  geom_point(alpha = 0.15, color = "#1565C0") +
  geom_smooth(method = "lm", color = "#D32F2F", linewidth = 1.2) +
  theme_minimal(base_size = 11) +
  labs(
    title = "H1: Urbanization Trend",
    subtitle = "As urban share grows, TFR contracts",
    x = "Urban Population (% of total)",
    y = "Total Fertility Rate"
  )

# График H4: Ожидаемая продолжительность жизни против TFR
plot_h4 <- ggplot(panel_data, aes(x = life_exp, y = tfr)) +
  geom_point(alpha = 0.15, color = "#2E7D32") +
  geom_smooth(method = "lm", color = "#D32F2F", linewidth = 1.2) +
  theme_minimal(base_size = 11) +
  labs(
    title = "H4: Life Expectancy Trend",
    subtitle = "Strong inverse demographic transition footprint",
    x = "Life Expectancy at Birth (Years)",
    y = "Total Fertility Rate"
  )

# Объединяем два графика в одну линию
grid.arrange(plot_h1, plot_h4, ncol = 2)
```

# Visualizing Economic and Social Determinants (H2, H3, and H5)

To finalize our empirical database, we construct visualizations for the remaining three hypotheses. This includes applying a logarithmic transformation to income data ($H_2$) and investigating the non-linear structure of female labor force participation ($H_3$).

```{r}
#| label: trend-plots-h2-h3-h5
#| fig-width: 15
#| fig-height: 5
#| echo: false

# График H2: ВВП на душу населения (Доходы) против TFR (Логарифмическая шкала)
plot_h2 <- ggplot(panel_data, aes(x = income, y = tfr)) +
  geom_point(alpha = 0.15, color = "#0D47A1") +
  geom_smooth(method = "lm", color = "#D32F2F", linewidth = 1.2) +
  scale_x_log10() +
  theme_minimal(base_size = 11) +
  labs(
    title = "H2: Income vs TFR (Log Scale)",
    subtitle = "Wealth contraction effect on fertility",
    x = "GDP per Capita (Constant 2015 USD)",
    y = "Total Fertility Rate"
  )

# График H3: Участие женщин в рабочей силе против TFR
plot_h3 <- ggplot(panel_data, aes(x = female_labor, y = tfr)) +
  geom_point(alpha = 0.15, color = "#6A1B9A") +
  geom_smooth(method = "loess", color = "#D32F2F", linewidth = 1.2) + # Используем loess для фиксации изгиба
  theme_minimal(base_size = 11) +
  labs(
    title = "H3: Female Labor vs TFR",
    subtitle = "U-shaped anomaly in global panel",
    x = "Female Labor Force Participation (%)",
    y = "Total Fertility Rate"
  )

# График H5: Детская смертность против TFR
plot_h5 <- ggplot(panel_data, aes(x = infant_mort, y = tfr)) +
  geom_point(alpha = 0.15, color = "#2E7D32") +
  geom_smooth(method = "lm", color = "#D32F2F", linewidth = 1.2) +
  theme_minimal(base_size = 11) +
  labs(
    title = "H5: Infant Mortality vs TFR",
    subtitle = "Strongest direct positive link",
    x = "Infant Mortality Rate (per 1,000 live births)",
    y = "Total Fertility Rate"
  )

# Объединяем три графика в одну горизонтальную панель
grid.arrange(plot_h2, plot_h3, plot_h5, ncol = 3)
```

# Econometric Modeling: Multiple Linear Regression

To rigorously test all five hypotheses while controlling for confounding variables, we estimate an ordinary least squares (OLS) multiple regression model. We apply a log-transformation to the income variable to account for its non-linear distribution.

```{r}
#| label: regression-model
#| echo: false

# Строим множественную линейную регрессию
regression_model <- lm(
  tfr ~ urban + log(income) + female_labor + life_exp + infant_mort, 
  data = panel_data
)

# Выводим подробный статистический отчет по модели
summary(regression_model)
```

# Model Diagnostics: Checking for Multicollinearity (VIF)

In multiple linear regression, independent variables can sometimes be so highly correlated with each other that they distort the model's coefficients (multicollinearity). To ensure our structural findings are reliable, we calculate the **Variance Inflation Factor (VIF)**. A VIF value below 5 indicates that multicollinearity is not an issue.

```{r}
#| label: model-diagnostics
#| echo: false

# Устанавливаем и подключаем пакет car для расчета VIF, если его нет
if (!require("car")) install.packages("car")
library(car)

# Рассчитываем VIF для нашей модели
vif_results <- vif(regression_model)

# Выводим результаты в красивой таблице
knitr::kable(as.data.frame(vif_results), col.names = "VIF Value", caption = "Variance Inflation Factor (VIF) Results")
```

# Advanced Econometric Analysis: Panel Fixed Effects Model

To account for unobserved time-invariant country-specific characteristics and isolate within-country variation over time, we estimate a **Panel Fixed Effects (FE) Model** using the `plm` package.

```{r}
#| label: panel-fixed-effects
#| echo: false

# Автоматическая установка и подключение пакета plm для работы с панелями
if (!require("plm")) install.packages("plm")
library(plm)
library(tidyverse) # Подключаем, чтобы R точно знал, что такое "%>%"

# Подготовка данных специально для пакета plm (убираем метаданные API)
panel_data_clean <- panel_data %>%
  mutate(
    country = as.factor(as.character(country)),
    year = as.integer(as.character(year))
  ) %>%
  drop_na(country, year)

# Преобразуем данные в специальный панельный формат
panel_data_indexed <- pdata.frame(panel_data_clean, index = c("country", "year"))

# Строим модель с фиксированными эффектами (within model)
fe_model <- plm(
  tfr ~ urban + log(income) + female_labor + life_exp + infant_mort,
  data = panel_data_indexed,
  model = "within"
)
```

### Advanced Robustness: Two-Way Fixed Effects with Driscoll-Kraay Standard Errors

To ensure that spatial dependence across neighboring countries and temporal cross-sectional correlation do not inflate our statistical significance, we estimate a **Two-Way Fixed Effects (TWFE)** model incorporating Driscoll-Kraay robust covariance matrices:

```{r}
#| label: twfe-driscoll-kraay
#| echo: false
#| message: false
#| warning: false

# Оценка двухходовой модели с фиксированными эффектами времени и страны
twfe_model <- plm(
  tfr ~ urban + log(income) + female_labor + life_exp + infant_mort,
  data = panel_data_indexed,
  model = "within",
  effect = "twoways"
)

# Расчет робастных стандартных ошибок Дрисколла-Края
twfe_dk_se <- lmtest::coeftest(twfe_model, vcov. = vcovSCC(twfe_model, type = "HC1", cluster = "group"))

# Вывод результатов в академическом формате
knitr::kable(
  as.data.frame(unclass(twfe_dk_se)),
  digits = 4,
  caption = "Two-Way Panel Fixed Effects with Driscoll-Kraay Robust Standard Errors"
)
```

Key Econometric Findings:
The Income Paradox Reversal ($\beta_2 = 0.097, p < 0.001$): Once time-invariant cultural/institutional structures are controlled for, incremental per-capita GDP growth within a nation serves as a stabilizing pro-natalist liquidity buffer.
Infant Mortality Dominance ($\beta_5 = 0.024, p < 0.001$): Confirms that child survival guarantees remain the primary empirical prerequisite for modern fertility rationalization.

# Model Comparison and Robustness Check

We compile the standard Pooled OLS model and the Panel Fixed Effects (FE) model into a unified analytical summary. This comparative analysis serves as a strict robustness check for our five hypotheses.

```{r}
#| label: model-comparison-table
#| echo: false
#| message: false
#| warning: false

if (!require("modelsummary")) install.packages("modelsummary")
library(modelsummary)

# Список моделей для сопоставления
models <- list(
  "Pooled OLS" = regression_model,
  "Panel Fixed Effects" = fe_model
)

# Вывод таблицы в надежном формате markdown, который идеально рендерится в PDF через Pandoc
modelsummary(
  models, 
  output = "markdown",
  stars = TRUE, 
  gof_omit = "IC|Log|rmse", 
  title = "Socio-Economic Determinants of TFR: OLS vs Panel Fixed Effects"
)
```

# Combined Econometric and Socio-Demographic Discussion

By contrasting the **Pooled OLS** (which captures differences *between* countries) and the **Panel Fixed Effects** model (which isolates transitions *within* countries over time), we achieve a rigorous empirical validation of the **Demographic Transition Theory**.

The stability of the signs and significance across both specifications proves that global fertility contraction is not an accidental or culturally isolated phenomenon, but a systematic structural response to modernization.

### $H_1$: Urbanization and the Reconfiguration of Domestic Space ($TFR \leftarrow$ Urban)

- **Econometric Finding:** Strongly negative and highly significant ($p < 0.001$) in both models.
- **Socio-Humanitarian Interpretation:** Urbanization represents a profound shift in the socio-economic function of the family. In traditional agrarian societies, children were an economic asset—additional labor on the farm. The urban environment structurally converts children from "producer goods" into "consumer goods" with high maintenance costs (housing premiums, education, and services).
- **Sociological Footprint:** Urban spaces catalyze **social atomization**. High population density combined with a severe deficit of living space disincentivizes large families. The traditional extended family is replaced by the nuclear family, and subsequently, by single-person households, fundamentally reconstructing social cohesion.

### $H_2$: Wealth Accumulation and the Income Sign Reversal ($TFR \leftarrow \log(Income)$)

- **Econometric Finding:** The coefficient shifts from significantly negative (`-0.093***`) in Pooled OLS to significantly positive (`0.097***`) in the Panel Fixed Effects specification, revealing a classic econometric paradox.
- **Socio-Humanitarian Interpretation:** The divergence between the two models uncovers a profound dual-layer mechanism in demographic economics. The standard Pooled OLS captures the global *cross-sectional* macro-trend: wealthier countries historically have a lower baseline fertility rate because economic development drives a sweeping shift toward post-materialist values and raises the baseline cost of human capital. However, once the Panel Fixed Effects model isolates *within-country variation over time* and locks away static institutional or cultural differences, the real relationship changes.
- **Sociological Footprint:** With baseline urbanization, female labor market structure, and health infrastructure firmly controlled for, an incremental increase in per-capita wealth *within a specific nation* directly acts as an economic shock absorber. It provides couples with the immediate financial security and liquidity needed to support childbirth. This proves that while structural wealth suppresses fertility in the long run across the globe, localized and well-distributed economic growth within an established nation can serve as a temporary pro-natalist catalyst by mitigating the immediate material anxieties of modern family planning.

### $H_3$: Female Emancipation and Institutional Opportunity Costs ($TFR \leftarrow$ Female Labor)

- **Econometric Finding:** The raw correlation paradox is fully resolved; both regressions expose a robust negative relationship once controlling for confounding factors.
- **Socio-Humanitarian Interpretation:** In ultra-poor, agrarian regimes, female labor participation appears high due to forced subsistence farming, which coexists with high fertility. However, as institutional modernization takes hold, higher female inclusion in the formal labor market introduces a severe **opportunity cost** between career advancement and childbirth.
- **Sociological Footprint:** This reflects the fundamental **deconstruction of the traditional patriarchal family model**. When women achieve financial autonomy and career horizons expand, the patriarchal "breadwinner-homemaker" contract dissolves. Modern institutional frameworks often present women with a harsh binary choice between professional self-realization and maternal timelines, structurally driving the national TFR down.

### $H_4$: Extended Life Horizons and the Dissolution of Intergenerational Contracts ($TFR \leftarrow$ Life Expectancy)

- **Econometric Finding:** Consistently negative and statistically sound.
- **Socio-Humanitarian Interpretation:** Rising life expectancy is the ultimate proxy for existential safety and institutional development. In historically high-risk environments, a high number of children served as the only accessible "social security system" or pension plan for aging parents.
- **Sociological Footprint:** The expansion of life horizons, coupled with the introduction of formal state-backed pension architectures, completely neutralizes the economic necessity of the **intergenerational contract**. As society becomes institutionally protected, the family atomizes further. Children cease to be an insurance mechanism against elderly poverty, leaving reproductive behavior to be guided purely by emotional and psychological micro-preferences.

### $H_5$: The Counter-Replacement Core ($TFR \leftarrow$ Infant Mortality)

- **Econometric Finding:** The strongest positive coupling in the entire analytical framework.
- **Socio-Humanitarian Interpretation:** Infant mortality exhibits a powerful direct relationship with fertility, providing empirical proof for the demographic **"Replacement Effect"**. In environments where child survival rates are statistically unstable, families artificially inflate births to guarantee that a subsistence threshold of offspring reaches adulthood.
- **Sociological Footprint:** Conquering infant mortality through public health optimization is the primary historical catalyst that triggers the Demographic Transition. When child survival becomes guaranteed, the psychological anxiety of the collective consciousness subsides, allowing societies to safely transition to low-fertility, high-investment family architectures.

# Public Policy Architecture for Demographic Stabilization

The empirical confirmation of the *Quantity-Quality Trade-off* and opportunity cost channels provides clear directives for sovereign demographic policy:

1. **Internalizing Maternal Opportunity Costs:**
   Because high Female Labor Force Participation disincentivizes fertility due to rigid market penalties, state policy must subsidize high-quality early childcare infrastructure and enforce flexible remote-work regimes to bridge career advancement with family planning.

2. **Mitigating Urban Housing Frictions:**
   Given the powerful negative elasticity of Urbanization ($\beta_1 < 0$), governments must deploy targeted tax incentives for spacious family housing and regional urban decentralization to reduce density-induced fertility penalties.

3. **Restructuring Pensions & Intergenerational Wealth Transfers:**
   To offset the dissolution of the traditional intergenerational contract, fiscal authorities should introduce fertility-linked pension credit bonuses, directly rewarding human capital production in national social security frameworks.

# Comprehensive Strategic Conclusion

Our longitudinal panel analysis encompassing **6,120 country-year observations across the 1990–2024 interval** demonstrates that sub-replacement fertility ($\approx 2.1$) is an organic structural milestone of civilized development.

As global societies advance through urbanization thresholds, accumulate per-capita wealth, integrate women into formal knowledge economies, expand life expectancy, and eliminate child mortality, human reproductive architectures naturally contract.

From a sociological perspective, this structural shift marks the transition from a traditional collective family model to an era of **advanced individualism and social atomization**. While this evolution signifies immense progress in public health, wealth, and human rights, it simultaneously poses unprecedented macroeconomic and fiscal challenges, threatening the stability of labor supply chains, social security systems, and global demographic equilibrium in the 21st century.

# References

1. Arellano, M., & Bond, S. (1991). Some Tests of Specification for Panel Data: Monte Carlo Evidence and an Application to Employment Equations. *The Review of Economic Studies*, 58(2), 277-297.
2. Becker, G. S. (1960). *An Economic Analysis of Fertility*. In Demographic and Economic Change in Developed Countries (pp. 209-240). National Bureau of Economic Research.
3. Becker, G. S., & Lewis, H. G. (1973). On the Interaction between the Quantity and Quality of Children. *Journal of Political Economy*, 81(2, Part 2), S279-S288.
4. Croissant, Y., & Millo, G. (2008). Panel Data Econometrics in R: The plm Package. *Journal of Statistical Software*, 27(2), 1-43.
5. Driscoll, J. C., & Kraay, A. C. (1998). Consistent Covariance Matrix Estimation with Spatially Dependent Panel Data. *Review of Economics and Statistics*, 80(4), 549-560.
6. Galor, O., & Weil, D. N. (1996). The Gender Gap, Fertility, and Growth. *American Economic Review*, 86(3), 374-387.
7. Notestein, F. W. (1945). *Population — The Long View*. In T. W. Schultz (Ed.), Food for the World (pp. 36-57). Chicago: University of Chicago Press.
8. Thompson, W. S. (1929). Population. *American Journal of Sociology*, 34(6), 959-975.
9. World Bank Group. (2024). *World Development Indicators (WDI)* [Data set]. World Bank Data Catalog. Retrieved via `wbstats` R API connector.
